Sequential Composition for Relaxed Memory: Pomsets with Predicate Transformers

### 2021-04-14 paper/intro.tex
related

In this paper, we show how pomsets with
preconditions~\cite{DBLP:journals/pacmpl/JagadeesanJR20} can be combined with
predicate transformers~\cite{DBLP:journals/cacm/Dijkstra75} to create a
compositional semantics for sequential composition.
\begin{itemize}
\item \S\ref{sec:model} presents the basic model, with few features
  required of the logic of preconditions, but a resulting lack of fidelity
  to exiting models,
\item \S\ref{sec:q} adds a model of \emph{quiescence} to the logic,
  required to model coherence (accessing $x$ has a precondition that $x$ is quiescent)
  and synchronization (a releasing write requires all locations to be quiescent),
\item \S\ref{sec:arm} adds the features required for efficient compilation
  to modern architectures: downgrading some synchronized accesses to relaxed,
  and removing read-read dependencies, and
\item \S\ref{sec:complications} show how to address common litmus tests.
\end{itemize}
% Acknowledgements go here, once we're not double-blinded.
The definitions in this paper have been formalized in Agda.

Because it is closely related, we expect that the memory-model results of
\cite{DBLP:journals/pacmpl/JagadeesanJR20} apply to our model, including
compositional reasoning for temporal safety properties and {local} \drfsc{}
as in \cite{Dolan:2018:BDR:3192366.3192421,DBLP:conf/ppopp/DongolJR19}.
In \textsection\ref{sec:arm}, we provide an alternative proof strategy for
efficient compilation to \armeight{}, which improves upon that of
\cite{DBLP:journals/pacmpl/JagadeesanJR20} by using a recent alternative
characterization of \armeight{}.

@@

Prominent solutions to the problem of dependency calculation include:
\begin{itemize}

\item \emph{syntactic} methods used in hardware models
  such as ARM or x86-TSO \cite{alglave},
\item \emph{speculative execution} methods (which give a semantics based on multiple executions
  of the same program) such as the Java Memory Model~\cite{Manson:2005:JMM:1047659.1040336} 
  and related models \cite{Jagadeesan:2010:GOS:2175486.2175503, DBLP:conf/popl/KangHLVD17, DBLP:journals/pacmpl/ChakrabortyV19},
  % the speculative operational semantics of~\cite{Jagadeesan:2010:GOS:2175486.2175503},
  % the promising semantics of~\cite{DBLP:conf/popl/KangHLVD17},
  % or the event structures semantics of~\cite{DBLP:journals/pacmpl/ChakrabortyV19},
\item \emph{rewriting} methods, which give an operational model
  up to syntactic rewrites, such as~\cite{Pichon-Pharabod:2016:CSR:2837614.2837616}, and
\item \emph{logical} methods, such as the pomsets with preconditions
  model of~\cite{DBLP:journals/pacmpl/JagadeesanJR20}.

\end{itemize}
\begin{itemize}
\item Models with strong dependencies that disallow compiler optimizations
  and efficient implementation on \armeight{}:
  \cite{Lamport:1979:MMC:1311099.1311750,
    DBLP:conf/pldi/LahavVKHD17,
    Dolan:2018:BDR:3192366.3192421,
    DBLP:conf/pldi/LahavVKHD17,
    DBLP:conf/lics/JeffreyR16,
    Boehm:2014:OGA:2618128.2618134,
    DBLP:journals/corr/abs-1804-04214}.
\item Models with weak dependencies that allow some forms of thin-air
  reasoning:
  \cite{Manson:2005:JMM:1047659.1040336,
    DBLP:conf/esop/JagadeesanPR10,
    DBLP:conf/popl/KangHLVD17,
    DBLP:journals/pacmpl/ChakrabortyV19}
\item Models that do not calculate dependencies:
  \cite{Batty:2011:MCC:1926385.1926394,
    DBLP:journals/pacmpl/WattRP19,
    DBLP:conf/pldi/WattPPBDFPG20,
    goMM}.
\end{itemize}
Like us, \citet{DBLP:journals/corr/abs-1804-04214} use pomsets to create a
compositional account of sequential composition.  However, their model
requires a fence after every read on \armeight{}, and uses syntactic
dependencies, thus invalidating many compiler optimizations.

\citet{DBLP:conf/snapl/MarinoMMNS15} argue that the ``silently shifting
semicolon'' is problematic for programmers, and thus concurrent languages
should guarantee sequential abstraction.  But processor and language
implementors balk at the prospect, due to significant costs.

>>>>

In this paper, we show how pomsets with
preconditions~\cite{DBLP:journals/pacmpl/JagadeesanJR20} can be combined with
predicate transformers~\cite{DBLP:journals/cacm/Dijkstra75} to create a
compositional semantics for sequential composition.
\begin{itemize}
\item \S\ref{sec:model} presents the basic model, which satisfies many
  desiderata, but not all.
\item \S\ref{sec:arm} shows two approaches for efficient implementation on
  Arm.  The first uses a suboptimal lowering for acquiring reads.  The second
  uses an optimal lowering, but requires changes to the definitions of
  parallel and sequential composition.
\item \S\ref{sec:variants} generalizes the basic semantics of read and write
  to validate compiler optimizations.
\end{itemize}
% Acknowledgements go here, once we're not double-blinded.
%The definitions in this paper have been formalized in Agda.
Because it is closely related, we expect that the memory-model results of
\cite{DBLP:journals/pacmpl/JagadeesanJR20} apply to our model, including
compositional reasoning for temporal safety properties and {local} \drfsc{}
as in \cite{Dolan:2018:BDR:3192366.3192421,DBLP:conf/ppopp/DongolJR19,promising-ldrf}.
% In \textsection\ref{sec:arm}, we provide an alternative proof strategy for
% efficient compilation to \armeight{}, which improves upon that of
% \cite{DBLP:journals/pacmpl/JagadeesanJR20} by using a recent alternative
% characterization of \armeight{}.

@@

\citet{DBLP:conf/snapl/MarinoMMNS15} argue that the ``silently shifting
semicolon'' sufficiently problematic for programmers that concurrent
languages should guarantee sequential abstraction, despite the performance
penalties.  In this paper, we have take the opposite approach.  We have
attempted to find the most intellectually tractable model that encompasses
all of the messiness of relaxed memory.

There are prior studies of relaxed memory that include sequential
composition.  \citet{DBLP:journals/corr/abs-1804-04214} define a semantics
using pomsets without preconditions. Instead, their model uses syntactic
dependencies, thus invalidating many compiler optimizations.  They also
require a fence after every relaxed read on \armeight{}.

\citet{DBLP:conf/esop/PaviottiCPWOB20} give a denotational semantics,
calculating dependencies using event structures rather than logic.  They give
the semantics of sequential composition in continuation passing style,
whereas we give it in direct style.

\citet{Pichon-Pharabod:2016:CSR:2837614.2837616} also use event structures to
calculate dependencies, combined with an operational semantics that
incorporates program transformations.  This approach seems to require
whole-program analysis.

Other studies of relaxed memory can be categorized by their approach to
dependency calculation.  Hardware models use syntactic dependencies
\cite{alglave}.  Many software models do not bother with dependencies at all
\cite{Batty:2011:MCC:1926385.1926394, DBLP:journals/pacmpl/WattRP19,
  DBLP:conf/pldi/WattPPBDFPG20, goMM}.  Others have strong dependencies that
disallow compiler optimizations and efficient implementation, typically
requiring fences for every relaxed read on Arm
\cite{Lamport:1979:MMC:1311099.1311750, DBLP:conf/pldi/LahavVKHD17,
  Dolan:2018:BDR:3192366.3192421, DBLP:conf/pldi/LahavVKHD17,
  DBLP:conf/lics/JeffreyR16, Boehm:2014:OGA:2618128.2618134,
  DBLP:journals/corr/abs-1804-04214}.

Many of the most prominent models are based on speculative execution
\cite{Manson:2005:JMM:1047659.1040336, DBLP:conf/esop/JagadeesanPR10,
  DBLP:conf/popl/KangHLVD17,DBLP:journals/pacmpl/ChakrabortyV19,DBLP:conf/pldi/LeeCPCHLV20,promising-ldrf}.
In their introduction, \citet{DBLP:journals/pacmpl/JagadeesanJR20} note that
these models fail to validate compositional reasoning of temporal
properties---see their examples OOTA4 and OOTA5 (from
\citep{DBLP:journals/toplas/Lochbihler13}).  The difference with our model
can be understood in terms of the valid program transformations.  The
speculative models allow reads to be introduced, with subsequent case
analysis on the value read---effectively, this can turn one read into two,
with different choices taken for the values read.  Our model invalidates this
transformation.  In return, our model enjoys compositionality for temporal
safety properties.
